
\hypertarget{data-analysis}{%
\section{Data Analysis}\label{data-analysis}}

\hypertarget{key-analysis-computing-challenges-at-the-hl-lhc}{%
\subsection{Key Analysis Computing Challenges at the
HL-LHC}\label{key-analysis-computing-challenges-at-the-hl-lhc}}

Examination of collision data is in essence the primary objective of the
experimental collaborations, coming at the end of the data
preparation/simulation and reconstruction chain. The stage of analysis,
starting in most cases from a standard data format generically referred
to as ``Analysis Object Data'' (AOD) containing reconstructed physics
object data, and producing physics results as the end product, poses
unique computing challenges. The scope of analysis data processing is
broad, encompassing the production of derived data formats as well as
end stage analysis actions such as producing histograms from those
intermediate datasets and visualising the final results. In the
following, attention is focused on the computing challenges related to
analysis for the ATLAS and CMS experiments.

Today, ROOT format AOD files and derived datasets take up the lion's
share of disk resources, filling approximately 80\% of the total data
volume for both ATLAS and CMS \cite{Ref1,Ref2}. To serve precision analyses
that require large event statistics, the HL-LHC will increase the
recorded event rate by an order of magnitude, compared with a projected
\textasciitilde 10\% annual growth of storage resources. The large pile-up in HL-LHC
collisions will compound the storage challenge by inflating the data
size per event. Reducing the storage footprint of data analysis is
therefore of paramount importance.

The effective CPU needs for end-stage analysis payloads are typically
orders of magnitude lower than those in simulation and reconstruction.
When scaled up to O(100) analyses per year, each processing the input
data dozens of times a year, the total CPU consumption at the HL-LHC is
projected to be around 10\% of total experimental computing \cite{Ref1,Ref2}.
Central production of analysis data formats may account for another
10-30\%. It is worth noting at this point that analysis data access
patterns tend to be more chaotic than preceding stages, which can
increase the effective resource needs by large factors.

More significant than the global storage and computational costs is the
job turnaround time, closely tied to the ``time to insight'', which is a
strong limitation on the speed of analysis publication. One significant
constraint is the need for full coverage of the input data, which is
inextricably linked to weaknesses in book-keeping tools that make it
difficult, if not impossible, to make meaningful studies on subsets of
the data. This full coverage requirement is clearly limited by computing
infrastructure load and downtime. Another challenge is the
multiplication of computing resource demands from the assessment of
systematic uncertainties, which involves processing many alternate
configurations and input data, inflating both the CPU load and storage
footprint. Improvement of community standards for metadata handling and
uncertainty handling will be very important for HL-LHC.

A last concern is that, while simulation and reconstruction code is
mostly optimised and developed under greater scrutiny, analysis code is
often produced with emphasis on quick results and less emphasis on code
quality or performance. Over time, this may result in inefficient,
under-documented code that exacerbates the disk and CPU shortfall and
poses a major difficulty when software is re-purposed for a new analysis.
Add to this a growing array of alternative toolkits from the data
science community, in particular for machine learning, and it is clear
that the entropy of the analysis code ecosystem has become a major
challenge in itself.

The following sections expand on the major challenges for analysis
software, including those identified above. An overview is given of
ongoing work on common tools that alleviate these problems, followed by
an outlook on R\&D prospects that would more significantly transform the
HL-LHC analysis model for major resource savings.

\hypertarget{analysis-processing-and-workflows-current-problems-and-solutions}{%
\subsection{Analysis Processing and Workflows: Current Problems and
Solutions}\label{analysis-processing-and-workflows-current-problems-and-solutions}}

To make data analysis tractable, AOD data are typically reduced both by
saving only relevant information for each event, which may require
transformations (object calibration, computation of derived variables)
beyond simply discarding information, and by skimming out only events
that are interesting. In post-reconstruction data formats, lossy
compression is also in use and being optimised. Coordination of the data
reduction phase is a key point for the organisation of data analysis
processing at HL-LHC.

Two approaches have been followed during LHC Run 2 by ATLAS and CMS:
ATLAS analysis trains \cite{Ref4} and the CMS mini-AOD, a highly reduced and
standardised event content. Analysis trains place software payloads
tailored for specific analyses (carriages) into a periodically scheduled
centralised execution, amortising data staging costs and sharing some
common processing steps \cite{Ref3}. Standardised event contents are instead
common reductions that satisfy the needs of the bulk of analyses,
avoiding duplication of commonly selected events in multiple datasets.
Neither of these approaches provides exhaustive coverage of analysis use
cases on the experiments and alternative data formats are needed for a
small but significant (10-20\%) fraction of analyses requiring custom
reconstruction that may be CPU-intensive. These exceptions to the rule
are expected to remain at the HL-LHC.

It is worth considering the viability of the arguably more flexible
option to have analysis trains run over the reconstruction output at the
HL-LHC. Extrapolating using a simple model, ATLAS finds that they would
produce 35 PB/year of data AOD and 213 PB/year of MC \cite{Ref1}. In
comparison, a reduced DAOD\_PHYSLITE format (10 kB/event) made from the AOD would
result in 0.5 PB/year for data and 2 PB/year for MC. Production of this
reduced format using a Data Carousel model \cite{Ref1} allows the much
larger AOD to reside on tape, providing significant savings in disk
storage. Experience on CMS shows that further reduction (down to
\textasciitilde2 kB/event for the nano-AOD) could be possible while
still supporting a majority of physics analyses \cite{Ref5}. File size
reductions also permit multiple copies to be held on disk for more
efficient parallel processing. It is clear that as many analyses as
possible need to be fed into this standardised analysis event-content
pipeline if analysis computing resources are to be kept under control.
Nevertheless, analysis trains or any similar means of synchronising data
access could be applied to this highly reduced format. It is worth
noting the LHCb analysis model already in Run 3 has to forego the luxury
of reprocessing Raw data. This constraint from sheer event rate is not
without merit as it greatly simplifies the data processing and analysis
models and removes one tier of offline data products.

Careful attention needs to be paid to the assessment of systematic
uncertainties, particularly with regard to event content
standardisation, to avoid creating many derived outputs that differ only
minimally. Analysis models that do not require all event information to
be present in a single file, instead leveraging ROOT ``friend trees'' or
columnar storage, may be a way to reduce this duplication both in the
case of uncertainty calculations and reconstruction augmentations, but
will require development of robust strategies for keeping the
augmentations in sync with the core events. It is noted that the ROOT
data format is quantifiably very efficient for the processing model in
HEP \cite{Ref16}, and any competitors would have to achieve a similar level
of performance.

Access to metadata, defined as cross-event information, is clearly a
weak point in the handling of the multiple analysis datasets originating
from the LHC. Metadata may include information such as book-keeping of
processed events, detector conditions and data quality, \emph{a
posteriori} measured scale factors, and software versioning. This
information is often scattered in multiple locations such as:
``in-file'' data, remote database resources, shared-area text or ROOT files,
TWiki pages or even in e-mail. The lack of proper metadata integration
and a coherent interface leaves analyses exposed to any problems in data
processing. This in turn results in data completeness demands that
simply will not work at the HL-LHC, where the huge datasets all but
guarantee that some fraction of the data will be unavailable at any
given point in time.

At the HL-LHC, as energy and luminosity conditions are quasi stable
across the years, datasets from multiple years with different conditions
will need to be analysed in a coherent way. The analysis software will
then need to be able to fetch and use the proper metadata automatically
while today's analysis software often requires dedicated configuration
and tuning for each data-taking period. Belle II have taken some steps
to address this by using the ``global tag'' concept commonly used in
reconstruction and applying it to analysis, allowing users to better
organise and store their analysis metadata. Nevertheless, until all
metadata is organised under the same global tag umbrella and accessed
through a coherent interface, which is extremely challenging, the
problem remains. Improvements here are necessary both to permit
efficient studies on partial datasets, and to reduce the risk of user
error in metadata access.

The growing complexity of analysis codes makes them extremely fragile
(i.e. bug-prone), hardly reusable, and unsuited for analysis
preservation needs. The current best efforts at analysis preservation
are based on a snapshot of the full analysis setup that can ensure the
possibility to re-run the code on the original dataset in a few years.
This may allow a future analyser to reproduce previous results, but will
likely not provide any understanding, for the future analyser, about
what the analysis was effectively doing. Another use case is to re-use
the analysis code on new data, be that the same dataset with improved
calibrations or additional data that will improve the precision of the
measurement. This brings significant additional challenges, not least
related to metadata.

Nevertheless, these preservation efforts encourage better organisation
of the analysis as a workflow, and promote the use of version control
tools together with continuous integration, which offers a natural route
to improving analysis code quality. In addition to these code quality
measures, a longer-term solution for code complexity may be the adoption
of a Domain-Specific Language (DSL) \cite{Ref10}, discussed in more detail
later, a model that would have physicists write logical descriptions of
their analyses, leaving low-level code generation and hardware
optimisation to a backend. Thus analysis design could be quickly
understood and shared, sidestepping the usual dearth of documentation,
while simultaneously isolating analysis design from implementation and
hardware, providing a natural means of analysis preservation. An
interesting development in this direction is the REANA platform \cite{Ref21}
that supports multiple backends and uses a DSL to provide a framework to
run an analysis. Already a valuable resource as a tool for analysing
open data, it is interesting to ask what would happen if every analyst
first had to structure their analysis and capture their workflows before
they started submitting jobs.

\hypertarget{plans-and-recent-progress}{%
\subsection{Plans and Recent Progress}\label{plans-and-recent-progress}}

Most of the problems identified above were already identified in the
community white paper \cite{Ref17} from 2017 and since then some progress
was made prototyping new technologies to enable data analysis at the
HL-LHC.

A first branch of new technologies is that of efficient data analysis
processing in the last steps of the analysis, i.e. when event data needs
to be selected, derived and aggregated into histograms. Several
platforms developed by industry or data science have emerged in recent
years to quickly aggregate large datasets with parallel execution
(either on clusters of computers or simply on multi-core servers). Many
of those tools have been tested to understand the feasibility of usage
in the HEP context providing fresh ideas and new paradigms that have
stimulated development within the HEP toolkit.

In particular, the Python ecosystem, implementing many such solutions
(e.g. numpy, pandas dataframes), is emerging as a possible alternative
to HEP's traditional C++ environment. This is thanks largely to a
combination of its versatility as a language, allowing rapid
prototyping, and the ability to out-source compute-intensive work to
more performant languages like C and C++. This is similar to the
experimental software frameworks where Python is used as steering code
and C/C++ is used as an efficient backend implementation. The typical
Belle II analysis starts with Python using a custom DSL to specify
particle selections and algorithms implemented in C++ and define output
ntuples. The ntuples are typically analysed by the PyHEP toolkit
\cite{Ref6}, and packages from the wider data science community. Deeper
integration with external python tools is also particularly important
for enabling state-of-the-art machine learning frameworks such as
PyTorch and TensorFlow, whose relevance in analysis is likely to grow in
the long term. Meanwhile PyROOT, which allows the use of any C++ class
from Python, has the potential to provide a coherent analysis ecosystem
with an effortless integration of HEP specific tools written in C++ with
industry python big data toolkits.

The ROOT package, i.e. the foundation and workhorse for LHC analysis,
has been the focus of substantial development. On the data storage
front, the ROOT team demonstrated better data compression and
accelerated reading for a wide variety of data formats. Information
density drives compression and this can vary massively between
experiments and analyses. This new format called RNTuple \cite{Ref9, ROOT-2020-HL-LHC}, an
evolution of the TTree file format, shows robust and significant
performance improvements (1-5x faster), which could potentially save
significant storage space (10-20\%) for the HL-LHC. Another area of
progress, and potential consolidation in ROOT I/O, is lossy compression,
where the number of significant bits for a quantity may be far fewer
than a standard storage type's mantissa, enabling further savings in
storage space~\cite{ROOT-2020-HL-LHC}.

Inspired by data science tools, the event processing framework was
extended with DataFrame-like functionality (RDataFrame) implementing a
declarative approach to analysis processing in either Python or C++
(always executed in C++), which natively exploits multi-core
architectures \cite{Ref7} and could be extended to accelerators. Used from
Python, it allows to pre-process data in C++, and export to numpy.
Flexibility in interfacing RDataFrame to non-ROOT data formats has
allowed the ALICE collaboration to prototype its Run 3 analysis model on
this new technology. Growing use of Machine Learning was also
anticipated, and the TMVA toolkit has been improved with substantial
optimisations, more complex deep learning models and enhanced
interoperability with common data science tools \cite{Ref14,Ref15}  which is of
particular importance for inference.

A complementary approach provided in Python is the toolchain of uproot,
awkward arrays and coffea \cite{Ref8} for efficient columnar operations and
transformations of the data. These function as a lightweight end-stage
analysis environment that provides only the elements necessary for
pre-processing ROOT inputs and converting them into formats used by the
standard ML frameworks, typically numpy. Lightweight distribution using
package managers such as pip or conda allows for rapid setup and
extension with other Python tools. Good performance appears to have been
achieved within the scope of these projects, which has been kept
focused, but for the purposes of fair comparison with other existing or
emerging options, defining clear benchmarks for I/O and processing speed
would be beneficial.

A further feature linked to the growing use of Python is the use of
``notebook'' technology such as Jupyter. This permits quasi-interactive
exploration where the annotated history including formatted outputs and
graphics can be saved and shared with collaborators for reproduction and
adaptation. Although not a substitute for command line scripts in
well-tested and complex workflows, notebooks make an effective vessel
for software education and have been leveraged as a powerful frontend
for access to facilities including CERN's SWAN \cite{Ref20}.

\hypertarget{prospects-and-rd-needs}{%
\subsection{Prospects and R\&D needs}\label{prospects-and-rd-needs}}

As the complexity of analysis data handling and processing grows,
developing efficient and robust code using a steadily increasing number
of tools running on ever more heterogeneous hardware becomes more and
more difficult. In addition, analysis code is typically re-implemented by
tens of individuals, in different experiments, analysis teams or
institutes mostly performing the same kind of operations, i.e. data
reduction, plotting, variation of systematic uncertainty, fitting, etc.
It is clear that tools and frameworks for performing these repeated
operations need to provide very efficient interfaces to avoid analysis
code bloat. Declarative rather than imperative programming is becoming
visibly more prevalent both in and outside of the field as it naturally
provides efficient syntax.

DSLs are a generalisation of this declarative concept and have already
demonstrated their ability to simplify analysis code. Not only do %% Missing reference?
they allow users to express operations concisely, improving
comprehension and reusability, DSLs provide a natural layer of
insulation against hardware evolution as low-level, optimised code
generation for multiple different platforms and architectures is
delegated to the tool and framework experts. Prototypes of DSLs have
been developed in recent years \cite{Ref10} for the event processing and
histogram production parts of the analysis workflow, and even built into
experimental frameworks to perform high-level analysis \cite{Ref19}. Similar
efforts can also be investigated in the context of data interpretation
and statistical analysis and an interesting example here is the so
called ``combine tools'' developed by CMS and based on RooFit for the
Higgs discovery). Descriptions of a full analysis in terms of %% Missing reference?
workflows could then tie together the analysis at the top level \cite{Ref18,Ref21}. With some effort, the HEP community may be able to converge on
effective tools that provide a common solution, ideally across
experiments.

There is an obvious need for good integration with analysis backends,
whether these be local CPU, batch or grid resources. At some stage in
the analysis process, interactivity or fast turn-around is needed. While
typical grid task completion has a time scale of one day to one week,
fast turn-around exploration requires answers within a few seconds up to
perhaps a few hours in order to keep people productive. As analysis
tasks are often IO-intensive and not necessarily-CPU intensive, the
right trade off between CPU and high bandwidth storage should be
carefully studied. More detailed studies of resource usage and its
evolution, especially as pertains to the need for heterogeneous compute,
is needed to ensure the right resource balance is available in the
HL-LHC computing infrastructure. This will require more interaction
between the analysis and infrastructure communities.

As a proof of principle, an unoptimised Higgs discovery analysis has
been rerun on Google Cloud in a few minutes as the infrastructure was
able to quickly provide tens of thousands of CPUs with guaranteed
bandwidth to storage of 2 Gb/coresec \cite{Ref12}. Scaling this to
sustained, diverse analysis payloads will be far from trivial, so the
question becomes how we can compare different approaches to provisioning
analysis. Eight initial benchmark challenges have been defined to set
the scope that should be addressed by analysis systems \cite{Ref11}. What
remains to be demonstrated is the capability of such systems to scale to
full analysis complexity, including tasks such as the handling of
systematic uncertainties and data-driven background estimation, and
these are yet to be incorporated into the benchmarks.

In contrast to the activity on analysis languages and infrastructure,
metadata is a topic that is almost entirely neglected by the analysis
community, even while computing experts have long understood its
importance. That is likely due to the absence of a single coherent
stake-holder for and provider of analysis metadata. Event data comes in
a ROOT file, but already within a single experiment there are numerous
sources of metadata that operate on vastly different timescales and this
poses a significant challenge to standardisation. As analyses and
detectors grow in complexity, further divergence may amplify the
challenge, making it critical to begin addressing this problem promptly.
Nevertheless, the drive to capture a complete analysis must drag
metadata with it, and the gleeful adoption of the declarative paradigm
will make database-like interfaces seem less scary than they once were
and may finally allow a more complete integration with analysis code. I
have a dream.

One final comment in reviewing recent developments is worth noting.
Exposure of the HEP community to industry-standard data science tools
has undoubtedly been a force for good and a net gain for the community,
providing fresh ideas and new ways of working - RDataFrame is an
excellent example. It is clear that the larger and better-supported data
science community will continue to innovate and produce tools that HEP
will benefit from, both by using directly and by borrowing their ideas
for our own tools. Taking machine learning as an example, it is
extremely likely that industry will blaze a trail that HEP would do well
to follow and we should continue to build bridges to those tools.
Equally though, HEP will continue to have its own unique challenges and
it's less likely that ultra-performant machine learning inference will
be top of the data scientists wishlist. Given the challenges presented
by extreme-throughput analysis at the HL-LHC, the right balance must be
found between software development within the community and relying on
external toolkits.

Finally, attention must be given to ensuring the longevity of any
projects that the HEP community decides to adopt en masse. Whether by
specific institutes, funding schemes or other collective programmes,
responsibility for key software will need to be ensured in the long
term. The development of a secure support base will therefore need to be
considered for emerging projects that show significant promise.
